Emergent dual scaling of riverine biodiversity

### 2021-06-15 manuscript_pnas.Rmd
edits: main text

**Abstract:** A prevailing paradigm suggests that species richness increases with area in a decelerating way; this ubiquitous power-law scaling, the species-area relationship, has formed the foundation of many conservation strategies. In spatially complex ecosystems, however, the area may not be the sole dimension to scale biodiversity patterns because the scale-invariant complexity of fractal ecosystem structure may drive ecological dynamics in space. Here, we use theory and analysis of extensive fish community data from two distinct geographic regions to show that riverine biodiversity follows a robust scaling law along the two orthogonal dimensions of ecosystem size and complexity. In river networks, the recurrent merging of various tributaries forms fractal branching systems, where the prevalence of branching (ecosystem complexity) represents a macro-scale control of the ecosystem's habitat heterogeneity. In the meantime, ecosystem size dictates metacommunity size and total habitat diversity, two factors regulating biodiversity in nature. Our theory predicted that, regardless of simulated species' traits, larger and more branched 'complex' networks support greater species richness due to increased space and environmental heterogeneity. The relationships were linear on logarithmic axes, indicating power-law scaling by ecosystem size and complexity. In support of this theoretical prediction, the power laws have consistently emerged in riverine fish communities across the study regions (Hokkaido Island in Japan and the midwestern United States) despite hosting different fauna with distinct evolutionary histories. The emergence of dual scaling law may be a pervasive property of branching networks with important implications for biodiversity conservation.

@@

Here, we propose a unified framework of ecosystem size and complexity in scaling biodiversity patterns in rivers - a prime example of complex branching ecosystems. Individual streams and rivers flow through different landscapes with distinct geological and climatic backgrounds, serving as a spatial unit of unique in-stream environments [@brownDendriticNetworkStructure2010; @petersonModellingDendriticEcological2013; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018; @isaakSlowClimateVelocities2016; @mcguireNetworkAnalysisReveals2014; @abbottUnexpectedSpatialStability2018]. The recurrent merging of diverse tributaries ultimately forms a fluvial network with fractal branching patterns [@rodriguez-iturbeFractalRiverBasins2001]. As such, the complexity of branching structure, which we define here as the probability of branching per unit river distance [@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014], may represent a macro-scale control of the ecosystem's habitat heterogeneity (habitat diversity per unit area) [@grantLivingBranchesPopulation2007; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018]. Meanwhile, ecosystem size (watershed area) should determine the metacommunity size and total habitat diversity (area x heterogeneity), two factors that regulate biodiversity at the metacommunity level [@hubbellUnifiedNeutralTheory2001; @kallimanisHowDoesHabitat2008]. Hence, riverine biodiversity may manifest scaling laws along the two orthogonal dimensions of branching networks.

@@

First, we developed a theoretical framework synthesizing the influences of ecosystem size and complexity on biodiversity patterns in branching ecosystems. We depicted branching ecosystems as a spatial network of connected habitat patches which local communities inhabit [@yeakelSynchronisationStabilityRiver2014; @teruiMetapopulationStabilityBranching2018] (**Figure 1A and B**). Two factors determine mean environmental conditions at each habitat patch: (1) headwater environments (the most upstream habitat patch) and (2) local environmental noise. Environmental values at headwaters are drawn randomly from a normal distribution and propagate downstream with local environmental noise (i.e., a spatial autoregressive process with white noise). These environmental values recurrently 'mix' at confluences considering the relative size of joining tributaries (see **Materials and Methods**). Therefore, our network-generation procedure resembles natural processes of how branching river networks create diverse habitats in a metacommunity [@mcguireNetworkAnalysisReveals2014]. We then simulated metacommunity dynamics in the theoretical branching networks using a general metacommunity model [@thompsonProcessbasedMetacommunityFramework2020]. In this model, 50 species with different abiotic niches (optimum and width) disperse along a network and compete for resources with varied strengths in temporally dynamic environments (see **Materials and Methods** for model details).

Given the results of extensive sensitivity analysis (**Supporting Information**), we considered `r n_scenario` simulation scenarios comprising a combination of four landscape and eight ecological scenarios. We distinguished landscape scenarios by setting four combinations of environmental variation at headwaters ($\sigma_h =$ `r unique(dat_sim$sd_env_source)`) and the degree of local environmental noise ($\sigma_l =$ `r unique(dat_sim$sd_env_lon)`), both of which are defined as standard deviations of normal distributions. When $\sigma_h > \sigma_l$, branching produces greater habitat heterogeneity because headwaters are the primary source of environmental variation (**Figure 1A**). This landscape scenario reproduces natural patterns of habitat heterogeneity, in which environmental conditions differ greatly among tributaries but are highly correlated within a tributary[@petersonModellingDendriticEcological2013; @mcguireNetworkAnalysisReveals2014]. In the meantime, when $\sigma_h \le \sigma_l$, local environmental noise masks environmental variation among tributaries, leading to limited influences of branching on habitat heterogeneity (**Figure 1B**). This scenario may reflect human-modified landscapes where the physical or biological distinctiveness of tributaries has been compromised due to human activities [@poffHomogenizationRegionalRiver2007; @teruiMetapopulationStabilityBranching2018]. Thus, the inequality between $\sigma_h$ and $\sigma_l$ creates contrasting patterns of habitat heterogeneity within a network. We also considered eight ecological scenarios that differ in dispersal distance (controlled by the rate parameter of an exponential dispersal kernel $\theta$), dispersal probability $p_{d}$, and the maximum value of interspecific competition strength $\alpha_{max}$ (see **Materials and Methods**). These simulation scenarios were capable of reproducing common spatial patterns of local species richness in rivers, corroborating the appropriateness of our choice in parameter combinations (**Supporting Information**). Under each simulation scenario, we simulated 1400 time steps of metacommunity dynamics (including 400 time steps of initialization and burn-in periods) in `r max(dat_sim$n_rep)` branching networks with the gradients of ecosystem size (the number of habitat patches: `r min(dat_sim$n_patch)` to `r max(dat_sim$n_patch)`) and complexity (branching probability: `r round(min(dat_sim$p_branch), 2)` to `r round(max(dat_sim$p_branch), 2)`).

Theory predicted that ecosystem size and complexity both increase $\gamma$ diversity under a realistic landscape scenario (**Figures 2 and 3**), where the environmental variation at headwaters $\sigma_h$ is greater than the degree of local environmental noise $\sigma_l$ (**Figure 1A**). These relationships followed power laws (i.e., linear in a log-log scale) and were consistent across ecological scenarios (compare panels in **Figures 2 and 3**). Importantly, the impact of ecosystem complexity was comparable to that of ecosystem size. Hence, regardless of ecological scenarios, ecosystem size and complexity are likely to be equally important in scaling $\gamma$ diversity.

However, dispersal processes affected mechanisms that underlie the positive effects of ecosystem size and complexity on $\gamma$ diversity (compare left and right columns in **Figures 2 and 3**). We observed a greater contribution of $\beta$ diversity (defined as $\frac{\gamma}{\alpha}$) to increased $\gamma$ diversity when dispersal limitation existed (i.e., species travel short distances). This pattern reflected significant spatial variation in species composition over the branching network and was likely driven by the local association of species' niche and abiotic environments (i.e., species sorting) [@leiboldMetacommunityConceptFramework2004; @thompsonProcessbasedMetacommunityFramework2020]. In contrast, when the dispersal limitation was relaxed (species travel long distances), a clear increase in $\alpha$ diversity underpinned the positive relationships between $\gamma$ diversity and ecosystem properties. The results agree with previous predictions that increased dispersal homogenizes community composition while enhancing local diversity through increased immigrants from suitable habitat patches (i.e., mass effects) [@leiboldMetacommunityConceptFramework2004; @economoSpeciesDiversityNeutral2008; @thompsonProcessbasedMetacommunityFramework2020]. Therefore, ecosystem size and complexity can increase $\gamma$ diversity via enhanced species sorting or mass effects, and which takes primacy is dependent on the degree of dispersal limitation. These patterns were qualitatively similar across different levels of dispersal probabilities (**Figures S5 and S12**). The strength of competition decreased the maximum of $\alpha$ diversity but did not change the scaling relationships with ecosystem properties (**Figures 2 and 3**). In summary, our theory highlights how apparently similar scaling in $\gamma$ diversity emerges through different ecological pathways.

Yet, influences of ecosystem size and complexity differed significantly in their dependence on landscape scenarios. Ecosystem size had positive effects on $\gamma$ diversity regardless of landscape scenarios, although the slopes were steeper with greater environmental variation (higher $\sigma_h$ and/or $\sigma_l$) (**Figures 2 and S5-S11**). This result is attributable to the fact that larger ecosystems can hold more individuals in a metacommunity [@hubbellUnifiedNeutralTheory2001]. In contrast, we observed limited or no influences of branching complexity when local environmental noise was equal to or exceeded environmental variation at headwaters ($\sigma_l \ge \sigma_h$; **Figures S13-S18**). Under this scenario, branching has a minor influence on the ecosystem's habitat heterogeneity because patch-level environmental variation is equivalent to or greater than environmental differences between tributaries (**Figure 1B**). Therefore, this landscape scenario eliminates the positive effect of branching complexity on $\gamma$ diversity. This theoretical prediction may not apply to pristine or semi-natural river networks where individual streams show distinct and spatially-correlated environmental conditions, including water temperature, water chemistry, and flow/sediment regimes [@petersonModellingDendriticEcological2013; @mooreEmergentStabilityLarge2015; @mcguireNetworkAnalysisReveals2014]. Instead, it may be most relevant to severely altered landscapes where localized human disturbance disrupts the environmental distinctiveness of branches through, for example, flow regulations by dams [@poffHomogenizationRegionalRiver2007]. Hence, our theory has important implications for riverine biodiversity conservation by pointing to the crucial role of habitat diversity produced by branching structure.

@@

Surprisingly, we found decisive support for the global model in explaining $\gamma$ diversity (Bayes factor = `r round(fit$gamma$bayes_factor, 1)`). The estimated $\gamma$ diversity linearly increased with increasing watershed area (ecosystem size) and branching probability (ecosystem complexity) on a log-log scale, and the estimated scaling exponents were consistent across the study regions (`r op(coef(fit$gamma$model)[2])` for watershed area and `r op(coef(fit$gamma$model)[3])` for branching probability; **Table 1 and Figure 4**). Thus, watershed-scale riverine biodiversity likely obeys a general scaling law along the two independent axes of ecosystem size and complexity. The regional consistency implies that ecological traits of constituent species have a minor effect on the scaling relationships, as theory predicts (**Figures 2 and 3**). It is also important to note that the effect of ecosystem complexity was striking in its magnitude. Average predictive comparisons [@gelmanAveragePredictiveComparisons2007] revealed that an expected increase of $\gamma$ diversity per 0.1 branching probability was `r op(ap_est$sim_estimate[2] * 0.1)` $\pm$ `r op(ap_est$sim_se[2] * 0.1)` species (**Materials and Methods** and **Supporting Information**); this level of increase in $\gamma$ diversity requires `r round(ap_est$sim_estimate[2] * 0.1 / ap_est$sim_estimate[1])` km^2^ in the watershed area (see **Table S7** for the estimated average predictive comparisons). Overall, the observed patterns were reasonable because numerous studies have shown that tributaries are the primary source of environmental heterogeneity in river networks [@brownDendriticNetworkStructure2010; @petersonModellingDendriticEcological2013; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018; @isaakSlowClimateVelocities2016; @mcguireNetworkAnalysisReveals2014; @abbottUnexpectedSpatialStability2018] - the sole theoretical assumption required for the dual scaling of $\gamma$ diversity to emerge.

Similarly, there were weak to moderate supports for the global models of $\alpha$ and $\beta$ diversity (Bayes factor = `r round(fit$alpha$bayes_factor, 1)` and `r round(fit$beta$bayes_factor, 1)`, respectively). In both regions, $\beta$ diversity responded significantly to ecosystem size and complexity while $\alpha$ diversity showing a weaker or a vague response to these ecosystem properties (**Table 1 and Figure 4**). In our simulations, this pattern has emerged under the scenarios with dispersal limitation, which elegantly matches the previous observations of stream fish movement. Field studies (e.g., mark-recapture) recurrently revealed the restricted movement of stream fish, typically limited to several hundreds of meters at an annual time scale [@radingerPatternsPredictorsFish2014]. The reciprocal agreement of theoretical and empirical patterns provides indirect but convincing evidence that dispersal limitation, which results in the increasing importance of species sorting process (**Figures 2 and 3**), plays a key role in driving the associations between $\gamma$ diversity and ecosystem properties in rivers.

The consistent effect of branching probability on $\gamma$ diversity across the study regions is noteworthy because many watersheds in the Midwest have been altered by agricultural land use (mean % agricultural land use: `r round(mean(wsd_subset[[2]]$frac_agri)*100)`% in the Midwest and `r round(mean(wsd_subset[[1]]$frac_agri)*100)`% in Hokkaido). If the intensive land use by humans impairs biological or physical distinctiveness among tributaries, theory predicts a weakened effect of branching probability on $\gamma$ diversity. However, $\gamma$ diversity increased significantly with increasing branching probability in this highly modified landscape, suggesting that tributaries still sustain unique environmental conditions to support high spatial variation in species composition. Indeed, $\beta$ diversity increased with increasing branching probability in both regions (**Figure 4**). It is conceivable that local geological and geomorphological differences, such as slope, aspect, and soil porosity, persist in human-dominated landscapes to maintain the diversity of in-stream processes among tributaries. The lack of land use effects (the fraction of agricultural land use and dam density) further corroborates our interpretation (**Table 1**). Although our analysis is correlative, the finding is encouraging because the branching complexity of river networks may serve as a natural defense system to human-induced environmental changes.

A potential criticism is that our results are driven by spurious correlations; however, this is very unlikely. We have statistically controlled the potential influences of climates and land use patterns (**Table 1**). Moreover, our empirical results were fairly consistent with theoretical predictions, which are free from any confounding factors. Thus, we are confident in our statistical inference.

@@

The excellent match between theoretical and empirical results provides strong evidence for the dual diversity scaling. In particular, the emerging complexity-diversity relationship points to several important avenues for riverine biodiversity conservation. First and foremost, there is now a clear need to explicitly consider the dimension of ecosystem complexity to achieve successful conservation. Human-induced habitat alterations, including flow regulation [@poffHomogenizationRegionalRiver2007], habitat fragmentation [@grillMappingWorldFreeflowing2019], and stream burial [@elmoreDisappearingHeadwatersPatterns2008], may compromise or restrict access to the diverse habitats that complex branching networks may support. Hence, it is imperative to recognize the role of branching complexity and minimize the homogenizing effects of human activities. Second, the complexity perspective may provide insights into the spatial planning of conservation efforts. Riverine reserves and local restoration are increasingly recognized as an effective management tool, and designing a spatial network of conservation sites in rivers is an area of active research [@koningNetworkGrassrootsReserves2020]. While large reserves or restoration sites are undoubtedly important, our results indicate the potential of coordinated networks of local conservation sites in protecting biodiversity. For example, synergies of multiple small reserves may emerge at the watershed scale when ecologically-distinct tributaries are involved in the design, as evidenced by the recent successful conservation of tropical fishes in Thailand's Salween basin [@koningNetworkGrassrootsReserves2020].

@@

$$
\begin{aligned}
\mu_{z,k} = \rho \mu_{z,k+1} + \epsilon_{k} &\text{(1)}
\end{aligned}
$$

>>>>

**Abstract:** A prevailing paradigm suggests that species richness increases with area in a decelerating way; this ubiquitous power-law scaling, the species-area relationship, has formed the foundation of many conservation strategies. In spatially complex ecosystems, however, the area may not be the sole dimension to scale biodiversity patterns because the scale-invariant complexity of fractal ecosystem structure may drive ecological dynamics in space. Here, we use theory and analysis of extensive fish community data from two distinct geographic regions to show that riverine biodiversity follows a robust scaling law along the two orthogonal dimensions of ecosystem size and complexity (i.e., the dual scaling law). In river networks, the recurrent merging of various tributaries forms fractal branching systems, where the prevalence of branching (ecosystem complexity) represents a macro-scale control of the ecosystem's habitat heterogeneity. In the meantime, ecosystem size dictates metacommunity size and total habitat diversity, two factors regulating biodiversity in nature. Our theory predicted that, regardless of simulated species' traits, larger and more branched 'complex' networks support greater species richness due to increased space and environmental heterogeneity. The relationships were linear on logarithmic axes, indicating power law scaling by ecosystem size and complexity. In support of this theoretical prediction, the power laws have consistently emerged in riverine fish communities across the study regions (Hokkaido Island in Japan and the Midwestern United States) despite hosting different fauna with distinct evolutionary histories. The emergence of dual scaling law may be a pervasive property of branching networks with important implications for biodiversity conservation.

@@

Here, we propose a unified framework of ecosystem size and complexity in scaling biodiversity patterns in rivers - a prime example of complex branching ecosystems. Individual streams and rivers flow through different landscapes with distinct geological and climatic backgrounds, serving as a spatial unit of unique in-stream environments [@brownDendriticNetworkStructure2010; @petersonModellingDendriticEcological2013; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018; @isaakSlowClimateVelocities2016; @mcguireNetworkAnalysisReveals2014; @abbottUnexpectedSpatialStability2018]. The recurrent merging of diverse tributaries ultimately forms a fluvial network with fractal branching patterns [@rodriguez-iturbeFractalRiverBasins2001]. As such, the complexity of branching structure, which we define here as the probability of branching per unit river distance [@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014], may represent a macro-scale control of the ecosystem's habitat heterogeneity (habitat diversity per unit area) [@grantLivingBranchesPopulation2007; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018]. Meanwhile, ecosystem size (watershed area) should determine the metacommunity size and total habitat diversity (area times heterogeneity), two factors that regulate biodiversity at the metacommunity level [@hubbellUnifiedNeutralTheory2001; @kallimanisHowDoesHabitat2008]. Hence, riverine biodiversity may manifest scaling laws along the two orthogonal dimensions of branching networks. We call this the dual scaling hypothesis of biodiversity.

@@

We developed a theoretical framework synthesizing the influences of ecosystem size and complexity on biodiversity patterns in branching ecosystems. We depicted branching ecosystems as a spatial network of connected habitat patches which local communities inhabit [@yeakelSynchronisationStabilityRiver2014; @teruiMetapopulationStabilityBranching2018] (**Figure 1A and B**). Two factors determine the mean environmental condition at each habitat patch $\mu_z$: (1) headwater environments (the most upstream habitat patch) and (2) local environmental noise. Environmental values at headwaters are drawn randomly from a normal distribution with a mean of zero and a standard deviation of $\sigma_h$ and propagate downstream. We modeled the downstream propagation as a spatial autoregressive process, in which local noise is added to the adjacent upstream environment as $\mu_{z,k} \sim Normal(\mu_{z,k+1}, \sigma_{l}^2)$ ($k$ is the network distance from the outlet patch; $k = 1$ at the outlet) . The environmental values recurrently 'mix' at confluences considering the relative size of joining tributaries (see **Materials and Methods**). Therefore, our network-generation procedure resembles natural processes of how branching river networks create diverse habitats in a metacommunity [@mcguireNetworkAnalysisReveals2014].

Metacommunity dynamics were then simulated in the theoretical branching networks using a general metacommunity model [@thompsonProcessbasedMetacommunityFramework2020]. In this model, the realized species' population growth is regulated by local abiotic environments, competition, and dispersal. Specifically, the number of individuals $N_{ix}(t + 1)$ for species $i$ at patch $x$ and time $t+1$ was given as:

$$
N_{ix}(t + 1) \sim Poisson(n_{ix}(t) + I_{ix}(t) - E_{ix}(t))
$$

where $n_{ix}(t)$ is the expected number of individuals given the local community dynamics at time $t$, $I_{ix}(t)$ the expected number of immigrants to patch $x$, and $E_{ix}(t)$ the expected number of emigrants from patch $x$. We assumed the Beverton-Holt equation to simulate local community dynamics [@bevertonDynamicsExploitedFish1957]:

$$
\begin{aligned}
n_{ix}(t) = \frac{N_{ix}(t)r_{ix}(t)}{1 + \frac{r_{0,i}-1}{K_{x}}\sum_{j=1}^S{b_{ij}N_{jx}(t)}}
\end{aligned}
$$

$r_{ix}(t)$ is the site- and time-specific reproductive rate of species $i$, $r_{0,i}$ the maximum reproductive rate, $K_{x}$ the site-specific carrying capacity, $b_{ij}$ the competition coefficient between species $i$ and $j$, and $S$ the number of species in a metacommunity ($S = 50$). Interspecific competition coefficients are random draws from a uniform distribution as $b_{ij, i \neq j} \sim Unif(0,b_{max})$ while keeping intraspecific competition coefficients constant ($b_{ii} = 1$). Here, the reproductive rate $r_{ix}(t)$ was related to local abiotic environments through a Gaussian function with species' niche optimum $\mu_i$ and width $\sigma_{niche,i}$. That is, $r_{ix}(t)$ decreases as temporally-variable abiotic environments (drawn randomly from the mean environment $\mu_{z,x}$) deviate from $\mu_i$ (see **Materials and Methods**). Species' reproductive performance has therefore an explicit linkage with local abiotic environments, which can be driven largely by branching structure. After local community dynamics, species emigrate from the local habitat patch with probability $p_d$ and disperse through the branching network following an exponential dispersal kernel with the rate parameter $\theta$ (the reciprocal $1/\theta$ represents an expected dispersal distance). These dispersal parameters affect the number of immigrants $I_{ix}(t)$ and emigrants $E_{ix}(t)$.

Prior to the main simulation analysis, we performed a systematic sensitivity analysis to identify simulation parameters that strongly influence the relationships between biodiversity metrics ($\alpha$, $\beta$, and $\gamma$ diversity) and ecosystem properties (size and complexity). Our analysis revealed that the following parameters were particularly influential: environmental variation at headwaters $\sigma_h$, local environmental noise $\sigma_l$, maximum value of interspecific competition coefficient $b_{max}$ , dispersal probability $p_d$, and dispersal rate parameter $\theta$ (**Supporting Information**). Therefore, we considered `r n_scenario` combinations of these parameters as primary simulation scenarios, which include a combination of four landscape and eight ecological scenarios.

The four landscape scenarios represent distinct patterns of spatial environmental heterogeneity, which were produced by different combinations of environmental variation at headwaters ($\sigma_h =$ `r unique(dat_sim$sd_env_source)`) and the degree of local environmental noise ($\sigma_l =$ `r unique(dat_sim$sd_env_lon)`). When $\sigma_h > \sigma_l$, branching produces greater habitat heterogeneity because headwaters are the primary source of environmental variation (**Figure 1A**). This landscape scenario reproduces natural patterns of habitat heterogeneity, in which environmental conditions differ greatly among tributaries but are highly correlated within a tributary [@petersonModellingDendriticEcological2013; @mcguireNetworkAnalysisReveals2014]. In the meantime, when $\sigma_h \le \sigma_l$, local environmental noise masks environmental variation among tributaries, leading to limited influences of branching on habitat heterogeneity (**Figure 1B**). This scenario may reflect human-modified landscapes, in which human activities may reduce the headwater diversity $\sigma_h$ by homogenizing in-stream processes (e.g., dams) [@poffHomogenizationRegionalRiver2007] and/or may increase the local environmental noise $\sigma_l$ through localized habitat degradation (e.g., removal of riparian forests) [@dala-corteThresholdsFreshwaterBiodiversity2020]. Thus, the inequality between $\sigma_h$ and $\sigma_l$ defines the contrasting scenarios of natural and human-dominated landscapes. The eight ecological scenarios distinguish ecological characteristics of constituent species in a metacommunity. Among the ecological scenarios, we changed dispersal distance ($\theta = 0.1, 1.0$), dispersal probability ($p_{d} = 0.01, 0.1$), and the maximum value of interspecific competition strength ($b_{max} = 0.75, 1.50$). This setup enables us to examine how ecological traits of constituent species mediate biodiversity relationships. The combinations of landscape and ecological scenarios were capable of reproducing common spatial patterns of local species richness in rivers, corroborating the appropriateness of our choice in parameter combinations (**Supporting Information**). Under each simulation scenario, we simulated 1400 time steps of metacommunity dynamics (including 400 time steps of initialization and burn-in periods) in `r max(dat_sim$n_rep)` branching networks with the gradients of ecosystem size (the number of habitat patches: `r min(dat_sim$n_patch)` to `r max(dat_sim$n_patch)`) and complexity (branching probability: `r round(min(dat_sim$p_branch), 2)` to `r round(max(dat_sim$p_branch), 2)`).

Our simulation predicted the emergence of dual scaling law under a realistic landscape scenario, where the environmental variation at headwaters $\sigma_h$ is greater than the degree of local environmental noise $\sigma_l$ (**Figure 1A**). Ecosystem size and complexity both increased $\gamma$ diversity (**Figures 2A and 3A**) with a signature of power law (i.e., linear in a log-log scale), and these relationships were consistent across ecological scenarios (compare panels in **Figures 2A and 3A**). Importantly, the impact of ecosystem complexity was comparable to that of ecosystem size. Hence, regardless of ecological scenarios, ecosystem size and complexity are likely to be equally important in scaling $\gamma$ diversity. Our results contrast with predictions from neutral models, in which the enhanced connectivity of highly branched networks reduced $\gamma$ diversity by facilitating the further expansion of an already widespread species [@muneepeerakulEvolvingBiodiversityPatterns2019]. Non-neutral components of our model should therefore have played a central role in generating the predicted patterns. Indeed, the robustness of the scaling law stemmed from the convergence of niche and dispersal assembly as described below.

Dispersal regulated community assembly processes that converged to the robust scaling relationships of $\gamma$ diversity (compare left and right columns in **Figures 2A and 3A**). We observed a greater contribution of $\beta$ diversity (defined as $\frac{\gamma}{\alpha}$) to increased $\gamma$ diversity when dispersal limitation existed (i.e., species travel short distances). This result reflected significant spatial variation in species composition over the branching network and was likely driven by the local association of species' niche and abiotic environments (i.e., species sorting) [@leiboldMetacommunityConceptFramework2004; @thompsonProcessbasedMetacommunityFramework2020]. In contrast, when the dispersal limitation was relaxed (species travel long distances), a clear increase in $\alpha$ diversity underpinned the positive relationships between $\gamma$ diversity and ecosystem properties. The results agree with previous predictions that increased dispersal homogenizes community composition while enhancing local diversity through increased immigrants from suitable habitat patches (i.e., mass effects) [@leiboldMetacommunityConceptFramework2004; @economoSpeciesDiversityNeutral2008; @thompsonProcessbasedMetacommunityFramework2020]. Therefore, ecosystem size and complexity can increase $\gamma$ diversity via enhanced species sorting or mass effects, and which takes primacy is dependent on the degree of dispersal limitation. These patterns were qualitatively similar across different levels of dispersal probabilities (**Figures S5 and S12**).

Interspecific competition had minimal influences on the underlying mechanisms (compare top and bottom rows in **Figures 2A and 3A**). Strong competition reduced $\alpha$ diversity as observed in former studies [@thompsonProcessbasedMetacommunityFramework2020]. Yet, it did not change the qualitative relationships between biodiversity and ecosystem properties (**Figures 2A and 3A**). Accordingly, macroecological patterns are expected to be robust to variation in species interaction networks. This result deserves attention given the long-lasting debate on how species interactions change the dynamics of ecological communities [@kadoyaPotentialOscillatorsKeystone2018; @ushioFluctuatingInteractionNetwork2018]. Collectively, our theoretical analysis highlights how apparently similar scaling in $\gamma$ diversity emerges through different ecological pathways by explicitly modeling niche and dispersal assembly processes.

It should be borne in mind that influences of ecosystem size and complexity differed significantly in their dependence on landscape scenarios. Ecosystem size had positive effects on $\gamma$ diversity regardless of landscape scenarios, although the slopes were steeper with greater environmental variation (higher $\sigma_h$ and/or $\sigma_l$) (**Figures 2A and S5-S11**). This result is attributable to the fact that larger ecosystems can hold more individuals in a metacommunity [@hubbellUnifiedNeutralTheory2001]. In contrast, we observed limited or no influences of branching complexity when local environmental noise was equal to or exceeded environmental variation at headwaters ($\sigma_h \le \sigma_l$; **Figures S13-S18**). Under this scenario, branching has a minor influence on the ecosystem's habitat heterogeneity because patch-level environmental variation is equivalent to or greater than environmental differences between tributaries (**Figure 1B**). Therefore, this landscape scenario eliminates the positive effect of branching complexity on $\gamma$ diversity. This theoretical prediction may not apply to pristine or semi-natural river networks where individual streams show distinct and spatially-correlated environmental conditions, including water temperature, water chemistry, and flow/sediment regimes [@petersonModellingDendriticEcological2013; @mooreEmergentStabilityLarge2015; @mcguireNetworkAnalysisReveals2014]. Instead, it may be most relevant to severely altered landscapes where human disturbance disrupts the environmental distinctiveness of branches through, for example, flow regulations by dams [@poffHomogenizationRegionalRiver2007]. Hence, our theory has important implications for riverine biodiversity conservation by pointing to the crucial role of habitat diversity produced by branching structure.

@@

Surprisingly, we found decisive support for the global model in explaining $\gamma$ diversity (Bayes factor = `r round(fit$gamma$bayes_factor, 1)`). The estimated $\gamma$ diversity linearly increased with increasing watershed area (ecosystem size) and branching probability (ecosystem complexity) on a log-log scale, and the estimated scaling exponents were consistent across the study regions (`r op(coef(fit$gamma$model)[2])` for watershed area and `r op(coef(fit$gamma$model)[3])` for branching probability; **Table 1 and Figures 2B and 3B**). Thus, watershed-scale riverine biodiversity likely obeys a general scaling law along the two independent axes of ecosystem size and complexity. The regional consistency implies that ecological traits of constituent species have a minor effect on the scaling relationships, as our theory predicts (**Figures 2A and 3A**). It is also important to note that the effect of ecosystem complexity was striking in its magnitude. Average predictive comparisons [@gelmanAveragePredictiveComparisons2007] revealed that an expected increase of $\gamma$ diversity per 0.1 branching probability was `r op(ap_est$sim_estimate[2] * 0.1)` $\pm$ `r op(ap_est$sim_se[2] * 0.1)` species (**Materials and Methods** and **Supporting Information**); this level of increase in $\gamma$ diversity requires `r round(ap_est$sim_estimate[2] * 0.1 / ap_est$sim_estimate[1])` km^2^ in the watershed area (see **Table S7** for the estimated average predictive comparisons). Overall, the observed patterns were reasonable because numerous studies have shown that tributaries are the primary source of environmental heterogeneity in river networks [@brownDendriticNetworkStructure2010; @petersonModellingDendriticEcological2013; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018; @isaakSlowClimateVelocities2016; @mcguireNetworkAnalysisReveals2014; @abbottUnexpectedSpatialStability2018] - the sole theoretical assumption required for the dual scaling of $\gamma$ diversity to emerge.

Similarly, there were weak to moderate supports for the global models of $\alpha$ and $\beta$ diversity (Bayes factor = `r round(fit$alpha$bayes_factor, 1)` and `r round(fit$beta$bayes_factor, 1)`, respectively). In both regions, $\beta$ diversity responded significantly to ecosystem size and complexity while $\alpha$ diversity showing a weaker or a vague response to these ecosystem properties (**Table 1 and Figures 2B and 3B**). In our simulations, this pattern has emerged under the scenarios with dispersal limitation, which elegantly matches the previous observations of stream fish movement. Field studies (e.g., mark-recapture) recurrently revealed the restricted movement of stream fish, typically limited to several hundreds of meters at an annual time scale [@radingerPatternsPredictorsFish2014]. The reciprocal agreement of theoretical and empirical patterns provides indirect but convincing evidence that dispersal limitation, which results in the increasing importance of species sorting process (**Figures 2 and 3**), plays a key role in driving the associations between $\gamma$ diversity and ecosystem properties in rivers.

The consistent effect of branching probability on $\gamma$ diversity across the study regions is noteworthy because many watersheds in the Midwest have been altered by agricultural land use (mean % agricultural land use: `r round(mean(wsd_subset[[2]]$frac_agri)*100)`% in the Midwest and `r round(mean(wsd_subset[[1]]$frac_agri)*100)`% in Hokkaido). If the intensive land use by humans impairs biological or physical distinctiveness among tributaries, theory predicts a weakened effect of branching probability on $\gamma$ diversity (**Figures S13-S18**). However, $\gamma$ diversity increased significantly with increasing branching probability in this highly modified landscape, suggesting that tributaries still sustain unique environmental conditions to support high spatial variation in species composition. Indeed, $\beta$ diversity increased with increasing branching probability in both regions (**Figure 4**). It is conceivable that local geological and geomorphological differences, such as slope, aspect, and soil porosity, persist in human-dominated landscapes to maintain the diversity of in-stream processes. The lack of land use effects (the fraction of agricultural land use and dam density) further corroborates our interpretation (**Table 1**). Although our analysis is correlative, the finding is encouraging because the branching complexity of river networks may serve as a natural defense system to human-induced environmental changes.

A potential criticism is that our results are driven by spurious correlations; however, this is very unlikely. We have statistically controlled the potential influences of climates and land use patterns (**Table 1**). Moreover, our empirical results were qualitatively consistent with theoretical predictions, which are free from any confounding factors. Thus, we are confident in our statistical inference.

@@

Recent advancement of spatial theories has sparked discussion of how the complex spatial structure of ecosystems drives macroecological patterns of diverse organisms. Several studies have explicitly considered the physical architecture of branching networks [@rinaldoRiverNetworksEcological2018] and have succeeded in reproducing some of the core biodiversity patterns (e.g., spatial distribution of local species richness) at coarse spatial resolutions (\~5000 km^2^) [@muneepeerakulNeutralMetacommunityModels2008a]. Yet, we have overlooked one of the fundamental rules in complex ecosystems, i.e., the dual scaling law. Here, the unified framework of ecosystem size and complexity provided strong evidence for the dual scaling law. Notably, this pattern has emerged irrespective of ecological contexts or geographic regions, emphasizing the generality of our findings. The dual scaling law of biodiversity may be an overlooked yet pervasive property of branching ecosystems.

In particular, the emerging complexity-diversity relationship points to several important avenues for riverine biodiversity conservation. First and foremost, there is now a clear need to explicitly consider the dimension of ecosystem complexity to achieve successful conservation. Human-induced habitat alterations, including flow regulation [@poffHomogenizationRegionalRiver2007], habitat fragmentation [@grillMappingWorldFreeflowing2019], and stream burial [@elmoreDisappearingHeadwatersPatterns2008], may compromise or restrict access to the diverse habitats that complex branching networks may support. Hence, it is imperative to recognize the role of branching complexity and minimize the homogenizing effects of human activities. Second, the complexity perspective may provide insights into the spatial planning of conservation efforts. Riverine reserves and local restoration are increasingly recognized as an effective management tool, and designing a spatial network of conservation sites in rivers is an area of active research [@koningNetworkGrassrootsReserves2020]. While large reserves or restoration sites are undoubtedly important, our results indicate the potential of coordinated networks of local conservation sites in protecting biodiversity. For example, synergies of multiple small reserves may emerge at the watershed scale when ecologically-distinct tributaries are involved in the design, as evidenced by the recent successful conservation of tropical fishes in Thailand's Salween basin [@koningNetworkGrassrootsReserves2020].

@@

$$
\begin{aligned}
\mu_{z,k} = \rho \mu_{z,k+1} + \epsilon_k &\text{(1)}
\end{aligned}
$$
